\vspace{1.0cm} \LARGE \bf Reassortment between influenza B lineages and the emergence of a co-adapted PB1-PB2-HA gene complex

### 2014-01-31 fluB.tex
More revisions.

Influenza B viruses are increasingly being recognized as major contributors to morbidity attributed to seasonal influenza. 
Currently circulating influenza B isolates are known to belong to two antigenically distinct lineages referred to as B/Victoria and B/Yamagata lineages. 
Frequent reassortment between the segments of these two lineages has been noted in the past, but the effects of these reassortments have not been investigated in much detail.
We investigate inter-lineage reassortments by modelling the lineages of all genomic segments on phylogenetic trees.
Our analyses indicate that of the 8 segments of influenza B viruses only PB1, PB2 and HA segments still have a Victoria and a Yamagata lineage and that currently circulating strains possess PB1, PB2 and HA segments derived entirely from one or the other lineage.
We argue that this is due to selection against reassortant viruses with mixed lineage PB1, PB2 and HA segments.
Given sufficient time we expect influenza B viruses to undergo sympatric speciation, if more segments get recruited by the already reassortment-isolated PB1-PB2-HA speciation island.

@@

We express the similarity of two trees by referring to the mean of the distribution of TMRCA differences ($\Delta$TMRCA) between pairs of tips in the two trees.
These parameters are estimated for each pairing of trees at each MCMC state and provides us with 95\% highest posterior density intervals for those parameters, thus giving us the ability to test specific hypotheses regarding similarities between the trees of different segments.
Our approach exploits the branch scaling used by BEAST \cite{drummond2012}, since the trees are scaled in absolute time and insensitive to variation in nucleotide substitution rates between segments, allowing for direct comparisons between TMRCAs in different trees.

If we are comparing TMRCAs of the same two tips in tree A and tree B we express it as 
\begin{equation}
\frac{[A_{i}:A_{i}']+[B_{i}:B_{i}']}{2\times [A_{i}:B_{i}]},
\frac{[A_{i}:A_{i}']+[B_{i}:B_{i}']}{2\times [A_{i}:B_{i}]},
\end{equation}
where $f(A_i,B_j)$ is the $\Delta$TMRCA of a pair of tips between the \textit{i}th tree derived from alignment A and the \textit{j}th tree derived from alignment B and $f(A_i,A_j)$ is the $\Delta$TMRCA between the same pair of tips between \textit{i}th tree derived from alignment A and \textit{j}th tree derived from an independent run of alignment A, to control for variability in tree topology stability over the course of the MCMC chain caused by differences in alignment lengths used to produce the trees.

@@

\subsection*{Estimating linkage disequilibrium across the influenza B genome}
We attempted to estimate linkage disequilibrium (LD) between amino acid sites across the longest proteins encoded by each segment of the influenza B virus genome.
We use the $\chi^{2}_{df}$ statistic \cite{zhao2005} for quantifying LD, as it is equal to the widely used \textit{r$^{2}$} LD statistic at biallelic loci, but also quantifies LD when there are more than two alleles per locus.
LD was estimated only at loci where each `allele' was represented by at least two isolates.
We ignored gaps in the alignment and did not consider them as polymorphisms.
However, there is at least one polymorphic indel within the HA protein, which is associated with the Vic-Yam split and not with incomplete sequences.

We also calculated mean LD for all pairs of segments to quantify the overall association between them.

\subsection*{Confirming findings with a larger dataset through lineage assignment}
The sequences of all 1433 available sequences of PB1, PB2, HA and NA segments from the same isolates sampled 1984-2013 were compiled from GISAID.
Neighbor-joining trees \cite{saitou1987} of PB1, PB2 and HA segments were made and each sequence assigned to a lineage based on grouping with either B/Victoria/2/87 or B/Yamagata/16/88 sequences.
Isolates which were not assigned entirely to either the Vic or the Yam lineage across PB1, PB2 and HA segments were extracted and identified as PB1-PB2-HA reassortants.

@@

The split of Vic and Yam lineages can be seen in all segments \cite{chen2008}.
Following the split of the two lineages, each segment could be assigned to either Vic or Yam lineage and inter-lineage reassortment events have yielded mixed-lineage genome constellations.
Some viruses with mixed-lineage genomes have become fixed in the influenza B virus population \textit{i.e.} became the `trunk' of the phylogenetic tree of a segment.

We find that Vic lineage PA, NP, NA and MP segments have gone extinct (\textit{i.e.} no Vic lineage sequences of those segments have been sampled in recent years see Figure \ref{lineageRatiosOverTime}), replaced entirely by the respective segments of Yam lineage in viruses bearing a Vic lineage HA.
Conversely, Vic lineage NS has become fixed in the influenza B virus population as well.
By looking at the most recent common ancestor of all lineages existing at yearly timeslices within the tree of each segment (see Figure \ref{tmrcaOT}), it is clear that PA, NP, NA, MP and NS segments have undergone periodic losses of diversity, whereby all sequences sampled more recently trace their ancestry to increasingly more recent common ancestors.

>>>>

Influenza B viruses are increasingly being recognized as major contributors to morbidity attributed to seasonal influenza. 
Currently circulating influenza B isolates are known to belong to two antigenically distinct lineages referred to as B/Victoria and B/Yamagata lineages. 
Frequent reassortment between the segments of these two lineages has been noted in the past, but the effects of these reassortments have not been investigated in much detail.
We investigate inter-lineage reassortments by comparing phylogenetic trees across genomic segments.
Our analyses indicate that of the 8 segments of influenza B viruses only PB1, PB2 and HA segments maintained separate Victoria and Yamagata lineages and that currently circulating strains possess PB1, PB2 and HA segments derived entirely from one or the other lineage; other segments have repeatedly reassorted between lineages thereby reducing genetic diversity.
We argue that this difference between segments is due to selection against reassortant viruses with mixed lineage PB1, PB2 and HA segments.
Given sufficient time we expect influenza B viruses to undergo sympatric speciation, if more segments get recruited by the already reassortment-isolated PB1-PB2-HA gene complex.

@@

We express the similarity of two trees by comparing to the mean of the distribution of TMRCA differences between pairs of tips in the two trees.
In comparing TMRCAs of trees A and B we calculate
\begin{equation}
\Delta(A_i, B_i) = \frac{f(A_i, A_i') + f(B_i, B_i')}{2 \, f(A_i, B_i)},
\end{equation}
where $f(A_i, B_i) = \frac{1}{n}\sum_{j=1}^n g(A_{ij}, B_{ij})$ and $g(A_{ij},B_{ij})$ is the difference in TMRCA of a pair of tips $j$, where the pair is drawn from the \textit{i}th posterior sample of tree A and the \textit{i}th posterior sample of tree B.
\tbc{Please correct if I misinterpreted what you were doing.}
Additionally, $f(A_i,A'_i)$ is calculated from the \textit{i}th posterior sample of tree A and \textit{i}th posterior sample of an independent analysis of tree A, to control for variability in tree topology stability over the course of the MCMC chain caused by differences in alignment lengths used to produce the trees.

These parameters are estimated for each pairing of trees at each MCMC state and provides us with 95\% highest posterior density intervals for those parameters, thus giving us the ability to test specific hypotheses regarding similarities between the trees of different segments.
Our approach exploits the branch scaling used by BEAST \cite{drummond2012}, since the trees are scaled in absolute time and insensitive to variation in nucleotide substitution rates between segments, allowing for direct comparisons between TMRCAs in different trees.

@@

\subsection*{Estimating linkage disequilibrium across the influenza B genome}
We estimated linkage disequilibrium (LD) between amino acid sites across the longest proteins encoded by each segment of the influenza B virus genome.
We used the $\chi^{2}_{df}$ \tbc{I couldn't download the article PDF, but is the $df$ really necessary?} statistic \cite{zhao2005} for quantifying LD, as it is equal to the widely used \textit{r$^{2}$} LD statistic at biallelic loci, but also quantifies LD when there are more than two alleles per locus.
LD was estimated only at loci where each nucleotide allele was represented by at least two isolates.
We ignored gaps in the alignment and did not consider them as polymorphisms.
We also calculated mean LD for all pairs of segments to quantify the overall association between them.

\subsection*{Confirming findings with a larger dataset through lineage assignment}
The sequences of all 1433 available sequences of PB1, PB2, HA and NA segments from the same isolates \tbc{`same' is confusing here, these are different isolates from those discussed previously?} sampled 1984-2013 were compiled from GISAID.
Neighbor-joining trees \cite{saitou1987} of PB1, PB2 and HA segments were made and each sequence assigned to a lineage based on grouping with either B/Victoria/2/87 or B/Yamagata/16/88 sequences.
Isolates which were not assigned entirely to either the Vic or the Yam lineage across PB1, PB2 and HA segments were extracted and identified as PB1-PB2-HA reassortants.

@@

The split into Vic and Yam lineages can be seen in all segments \cite{chen2008} \tbc{also reference supplemental tree figure}.
Following the split of the two lineages, each segment could be assigned to either Vic or Yam lineage and inter-lineage reassortment events have yielded mixed-lineage genome constellations.
Some viruses with mixed-lineage genomes have become fixed in the influenza B virus population \textit{i.e.} became the `trunk' of the phylogenetic tree of a segment.

We find that Vic lineage PA, NP, NA and MP segments have gone extinct (\textit{i.e.} no Vic lineage sequences of those segments have been sampled in recent years see Figure \ref{lineageRatiosOverTime}) \tbc{figures should not be referenced out of order, either make this figure 2 or remove this reference, or push this section back}, replaced entirely by the respective segments of Yam lineage in viruses bearing a Vic lineage HA.
Conversely, Vic lineage NS has become fixed in the influenza B virus population as well.
By looking at the most recent common ancestor of all lineages existing at yearly timeslices within the tree of each segment (see Figure \ref{tmrcaOT}) \tbc{again, restructure to fix out of order figure reference}, it is clear that PA, NP, NA, MP and NS segments have undergone periodic losses of diversity, whereby all sequences sampled more recently trace their ancestry to increasingly more recent common ancestors.

\tbc{This reads more like a beginning Discussion paragraph. These first Results paragraphs should be closer to the data.  Just describe what you show in figures.}
